---
title: "Applying Crossmaps"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Applying Crossmaps}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: knitr-opts
#| include: false

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
#| label: setup
#| message: false

library(xmap)
library(dplyr)
```

Once you have created and validated your `xmap_tbl` objects,
it's time to actually harmonise dataset values.
When transforming a dataset from one classification to another,
you have to make sure that you have rules for how to handle every single category in the source classification,
and that missing values are handled appropriately so you don't create silent errors with data loss.

This vignette covers the conditions `apply_xmap(.data, .xmap)` checks before transforming data, 
how to check them cheaply with `validate_apply_xmap()` or diagnose them in detail with
`diagnose_apply_xmap()`, and what to do if your crossmap doesn't
cover one or more of the source keys in your data.

We use `demo$simple_links`, a small `xcode -> alphacode` crossmap with a mix
of unit and fractional weights applied to `simple_data` as a self-contained running example:

```{r}
#| label: setup-data

simple_xmap <- demo$simple_links |>
  as_xmap_tbl(xcode, alphacode, weight)
simple_xmap

simple_data <- simple_links |>
  distinct(xcode) |>
  mutate(xcode_mass = 100)
simple_data
```

Here is a simple visualisation of the intended transformation in matrix form:

<!-- TODO: copy matrix visualisations from xmap talks -->

## Applying the transformation

The function `apply_xmap(.data, .xmap)` matches `.data`'s `keys_from` column against `.xmap$.from`,
multiplies each matched `values_from` value by its `.weight_by`, and sums the
results by `.to`:

```{r}
#| label: apply-basic

apply_xmap(
  simple_data,
  simple_xmap,
  values_from = xcode_mass,
  keys_from = xcode
)
```

Every `xcode_mass = 100` either passes through unchanged (unit weights, e.g.
`x1111 -> A1`) or splits proportionally across its `alphacode` targets (e.g.
`x2222`'s 100 splits 50/50 into `B2`/`B3`).

Before doing this arithmetic, `apply_xmap()` checks two conditions on
`.data` and aborts if either fails, rather than silently producing a wrong or
incomplete result: 

1. every `keys_from` key must have a matching link in
`.xmap$.from`, 
2. and no `values_from` column may hold a missing value.

`validate_apply_xmap()` checks the same two conditions and returns a single
`TRUE`/`FALSE`, without building any detail -- useful for a quick check
across many `.data`/`.xmap` pairs (e.g. inside a `dplyr::mutate()` over a
nested `country`/`year` collection, as in
`vignette("examine-compose-crossmaps")`) before applying any of them:

```{r}
#| label: validate-basic

validate_apply_xmap(
  simple_data,
  simple_xmap,
  values_from = xcode_mass,
  keys_from = xcode
)
```

If `validate_apply_xmap()` says something's wrong, you can use
`diagnose_apply_xmap()` to find out what and where. It checks the same two
conditions but also returns an `xmap_diagnosis` object with the offending rows attached.

<!-- TODO: IMAGE!!!! -->

## Diagnosing invalid transformation

### Discovering missing coverage

Every key in `.data$keys_from` must have a matching source key in `.xmap$.from`.
Without this, `apply_xmap()` has no way to transform a value it has no weights for.
Suppose `simple_xmap` is missing links for `x7777`:

```{r}
#| label: partial-xmap

partial_xmap <- simple_links |>
  filter(xcode != "x7777") |>
  as_xmap_tbl(xcode, alphacode, weight)
```

`diagnose_apply_xmap()` flags the uncovered key and attaches the affected
rows under `$details$not_covered`:

```{r}
#| label: diagnose-coverage

diagnose_apply_xmap(
  simple_data,
  partial_xmap,
  values_from = xcode_mass,
  keys_from = xcode
)
```

`apply_xmap()` itself aborts with a `coverage_error` on the same input,
rather than silently dropping `x7777`'s mass from the output:

```{r}
#| label: apply-coverage-error
#| error: true

apply_xmap(
  simple_data,
  partial_xmap,
  values_from = xcode_mass,
  keys_from = xcode
)
```

## Missing values

Missing values in your source data can lead to a number of implicit decisions in both the transformation of data and the interpretation of the transformed data. Although it is common to try and preserve missing values during recoding of labels, unless you only have 1-to-1 recodings, trying to split up or aggregate `NA` values often requires some implicit decision to treat the missing values as `0`.

For example, if a target `.to` category has input from three different `.from` categories, and one of those inputs is missing, the aggregated `.to` value could be something like `sum(NA, 100, 100)`. With `na.rm = FALSE`, the result is `NA`; with `na.rm = TRUE`, it's `200`. That `200` does match the *reported* total -- the sum of the numbers you can actually check, since the missing observation never had a checkable number to begin with -- but only because `na.rm = TRUE` silently treats it as if it contributed exactly `0`. That's an assumption, not a fact: the *true* total (whatever that observation's real, unrecorded value was) is unknown, and could easily not be `0`. A crossmap's mass-preservation guarantee (Definition 3.1's requirement that every value be a genuine $\mathbb{R}^+$ observation) is about that true total, not the reported one -- which is exactly why treating a missing value as `0` isn't something `apply_xmap()` does on your behalf.

From a transparency and reproducibility perspective, we think it is better to resolve missing values explicitly, before transforming, than to have `apply_xmap()` coerce them silently. To this end, `apply_xmap()` aborts on any missingness in `values_from`:

```{r}
#| label: na-data

na_data <- simple_data
na_data$xcode_mass[na_data$xcode == "x1111"] <- NA
na_data$xcode_mass[na_data$xcode == "x6666"] <- NA
```

`apply_xmap()` aborts with a `missing_mass_values` condition on the same
input:

```{r}
#| label: apply-missing-value-error
#| error: true

apply_xmap(
  na_data,
  simple_xmap,
  values_from = xcode_mass,
  keys_from = xcode
)
```

The fix is to remove or replace the missing value(s) before calling
`apply_xmap()` -- there is no `na.rm`-style option, since silently dropping
mass would break the total-preservation guarantee a valid crossmap provides.

`diagnose_apply_xmap()` flags this and attaches the affected rows under
`$details$missing_values`:

```{r}
#| label: diagnose-missing-value

diagnose_apply_xmap(
  na_data,
  simple_xmap,
  values_from = xcode_mass,
  keys_from = xcode
)
```

Currently, the strict failure on any `NA` values in the source data applies uniformly, regardless of how each key maps under
`.xmap`. This also disallows the 'unambiguous' preservation or propagation of missing values. For example, if we allowed for the pass through of `NA`s in cases without aggregation, `x1111`'s `NA` could pass straight through into `A1`, since `A1` has no other inputs. However, this requires some tiered logic if we want to make all implicit `0` coercions explicit. This is a current limitation of the package that we may relax in the future.

## Adding and removing keys

The two coverage failures above both arise from the same underlying
situation: `.xmap` and `.data` disagree about which keys exist. This happens
for real reasons -- classifications get revised, categories are born and
retired between one reporting period and the next -- so it's worth being
deliberate about where that disagreement gets resolved.

`apply_xmap()` deliberately does **not** try to resolve it for you. As the
crossmaps framework paper puts it (Section 6.2.3, ["A Unified Statistical And
Computational Framework For Ex-Post Harmonisation Of Aggregate
Statistics"](https://arxiv.org/abs/2406.14163)):

> The addition and removal of index keys between revisions of statistical
> classifications should also be handled outside of the crossmap transform
> rather than implicitly within the operation. (...) Rather than combine
> recoding and redistribution actions with removing or appending elements of
> category-indexed variables, we suggest removing any unwanted categories
> prior to applying the crossmap, and attaching any new categories after
> transforming the existing data.

The reasoning: a crossmap transform is defined to *preserve* the shared
total mass it redistributes. A source key with no target link, or a target
key with no source link, has no well-defined weight -- there's no principled
number to redistribute *to* or *from* it, only a modelling choice about
whether that mass should exist at all. Folding that choice into
`apply_xmap()` itself would mean the function sometimes creates or destroys
mass rather than only ever redistributing it, undermining the guarantee
`apply_xmap()`'s coverage and missing-value checks above exist to protect.

**Removing a retired source key.** Suppose `x7777` is retired in this
reporting period -- it no longer appears as a real observation. Rather than
leave a link for it in `.xmap` and hope it's simply never used, or strip it
from `.xmap` and risk a genuine, unretired `x7777` observation silently
erroring elsewhere, filter it out of `.data` before applying the transform:

```{r}
#| label: remove-retired-key

current_data <- simple_data |>
  filter(xcode != "x7777")

apply_xmap(
  current_data,
  simple_xmap,
  values_from = xcode_mass,
  keys_from = xcode
)
```

`simple_xmap` can safely keep its `x7777 -> D6` link -- `apply_xmap()` only
checks that every key in `.data` is covered by `.xmap`, not the reverse, so
an unused link is harmless. Only the data being transformed needs pruning.

**Adding a newly-born target key.** Suppose this period also introduces a
new category `F9`, reported directly with no corresponding `xcode` source
key to redistribute it from -- there is nothing in `simple_data` for a
crossmap to transform into it. Apply the transform to the data the crossmap
*does* cover, then attach the new category's value afterwards:

```{r}
#| label: add-new-key

transformed <- apply_xmap(
  simple_data,
  simple_xmap,
  values_from = xcode_mass,
  keys_from = xcode
)

new_category <- tibble::tibble(alphacode = "F9", xcode_mass = 40)

bind_rows(transformed, new_category)
```

In both cases, the crossmap transform itself stays a pure redistribution
step -- every `xcode_mass` going in is accounted for coming out, nothing
more and nothing less -- and the actual judgment call about which categories
exist this period is made explicitly, outside `apply_xmap()`, where it's
visible rather than implicit.
