---
title: "Applying Crossmaps"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Applying Crossmaps}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: knitr-opts
#| include: false

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
#| label: setup
#| message: false

library(xmap)
library(dplyr)
```

`vignette("extract-validate-existing")` and `vignette("examine-compose-crossmaps")`
are both about building and validating `xmap_tbl` objects themselves. This
vignette is about the other half of the workflow: `apply_xmap()`, which uses a
validated `xmap_tbl` to actually transform a dataset's values. It covers the
conditions `apply_xmap()` checks before transforming data, how to check them
cheaply with `validate_apply_xmap()` or diagnose them in detail with
`diagnose_apply_xmap()`, and how to handle keys that a crossmap doesn't
cover.

We use `demo$simple_links`, a small `xcode -> alphacode` crossmap with a mix
of unit and fractional weights, as a self-contained running example:

```{r}
#| label: setup-data

simple_links <- demo$simple_links
simple_links

simple_xmap <- simple_links |>
  as_xmap_tbl(xcode, alphacode, weight)

simple_data <- simple_links |>
  distinct(xcode) |>
  mutate(xcode_mass = 100)
simple_data
```

## Basic application

`apply_xmap()` matches `.data`'s `keys_from` column against `.xmap$.from`,
multiplies each matched `values_from` value by its `.weight_by`, and sums the
results by `.to`:

```{r}
#| label: apply-basic

apply_xmap(
  simple_data,
  simple_xmap,
  values_from = xcode_mass,
  keys_from = xcode
)
```

Every `xcode_mass = 100` either passes through unchanged (unit weights, e.g.
`x1111 -> A1`) or splits proportionally across its `alphacode` targets (e.g.
`x2222`'s 100 splits 50/50 into `B2`/`B3`).

Before doing this arithmetic, `apply_xmap()` checks two conditions on
`.data` and aborts if either fails, rather than silently producing a wrong or
incomplete result: every `keys_from` key must have a matching link in
`.xmap$.from`, and no `values_from` column may hold a missing value.

`validate_apply_xmap()` checks the same two conditions and returns a single
`TRUE`/`FALSE`, without building any detail -- useful for a quick check
across many `.data`/`.xmap` pairs (e.g. inside a `dplyr::mutate()` over a
nested `country`/`year` collection, as in
`vignette("examine-compose-crossmaps")`) before applying any of them:

```{r}
#| label: validate-basic

validate_apply_xmap(
  simple_data,
  simple_xmap,
  values_from = xcode_mass,
  keys_from = xcode
)
```

Once `validate_apply_xmap()` says something's wrong, reach for
`diagnose_apply_xmap()` to find out what and where. It checks the same two
conditions but returns an `xmap_diagnosis` object -- rather than a bare
logical -- with the offending rows attached, mirroring
`diagnose_as_xmap_tbl()`'s role for validating the crossmap itself.

## Coverage conditions

Every key in `.data$keys_from` must have a matching link in `.xmap$.from` --
`apply_xmap()` has no way to transform a value it has no weights for.
Suppose `simple_xmap` is missing links for `x7777`:

```{r}
#| label: partial-xmap

partial_xmap <- simple_links |>
  filter(xcode != "x7777") |>
  as_xmap_tbl(xcode, alphacode, weight)
```

`diagnose_apply_xmap()` flags the uncovered key and attaches the affected
rows under `$details$not_covered`:

```{r}
#| label: diagnose-coverage

diagnose_apply_xmap(
  simple_data,
  partial_xmap,
  values_from = xcode_mass,
  keys_from = xcode
)
```

`apply_xmap()` itself aborts with a `coverage_error` on the same input,
rather than silently dropping `x7777`'s mass from the output:

```{r}
#| label: apply-coverage-error
#| error: true

apply_xmap(
  simple_data,
  partial_xmap,
  values_from = xcode_mass,
  keys_from = xcode
)
```

## Missing values

Every value being transformed must be non-missing -- an `NA` split across
several `.to` targets would otherwise fan out silently, turning one missing
observation into several. Suppose `x2222`'s mass is unknown:

```{r}
#| label: na-data

na_data <- simple_data
na_data$xcode_mass[na_data$xcode == "x2222"] <- NA
```

`diagnose_apply_xmap()` flags this and attaches the affected rows under
`$details$missing_values`:

```{r}
#| label: diagnose-missing-value

diagnose_apply_xmap(
  na_data,
  simple_xmap,
  values_from = xcode_mass,
  keys_from = xcode
)
```

`apply_xmap()` aborts with a `missing_mass_values` condition on the same
input, since `x2222`'s missing mass would otherwise silently propagate `NA`
into both `B2` and `B3`:

```{r}
#| label: apply-missing-value-error
#| error: true

apply_xmap(
  na_data,
  simple_xmap,
  values_from = xcode_mass,
  keys_from = xcode
)
```

The fix is to remove or replace the missing value(s) before calling
`apply_xmap()` -- there is no `na.rm`-style option, since silently dropping
mass would break the total-preservation guarantee a valid crossmap provides.

## Adding and removing keys

The two coverage failures above both arise from the same underlying
situation: `.xmap` and `.data` disagree about which keys exist. This happens
for real reasons -- classifications get revised, categories are born and
retired between one reporting period and the next -- so it's worth being
deliberate about where that disagreement gets resolved.

`apply_xmap()` deliberately does **not** try to resolve it for you. As the
crossmaps framework paper puts it (Section 6.2.3, ["A Unified Statistical And
Computational Framework For Ex-Post Harmonisation Of Aggregate
Statistics"](https://arxiv.org/abs/2406.14163)):

> The addition and removal of index keys between revisions of statistical
> classifications should also be handled outside of the crossmap transform
> rather than implicitly within the operation. (...) Rather than combine
> recoding and redistribution actions with removing or appending elements of
> category-indexed variables, we suggest removing any unwanted categories
> prior to applying the crossmap, and attaching any new categories after
> transforming the existing data.

The reasoning: a crossmap transform is defined to *preserve* the shared
total mass it redistributes. A source key with no target link, or a target
key with no source link, has no well-defined weight -- there's no principled
number to redistribute *to* or *from* it, only a modelling choice about
whether that mass should exist at all. Folding that choice into
`apply_xmap()` itself would mean the function sometimes creates or destroys
mass rather than only ever redistributing it, undermining the guarantee
`apply_xmap()`'s coverage and missing-value checks above exist to protect.

**Removing a retired source key.** Suppose `x7777` is retired in this
reporting period -- it no longer appears as a real observation. Rather than
leave a link for it in `.xmap` and hope it's simply never used, or strip it
from `.xmap` and risk a genuine, unretired `x7777` observation silently
erroring elsewhere, filter it out of `.data` before applying the transform:

```{r}
#| label: remove-retired-key

current_data <- simple_data |>
  filter(xcode != "x7777")

apply_xmap(
  current_data,
  simple_xmap,
  values_from = xcode_mass,
  keys_from = xcode
)
```

`simple_xmap` can safely keep its `x7777 -> D6` link -- `apply_xmap()` only
checks that every key in `.data` is covered by `.xmap`, not the reverse, so
an unused link is harmless. Only the data being transformed needs pruning.

**Adding a newly-born target key.** Suppose this period also introduces a
new category `F9`, reported directly with no corresponding `xcode` source
key to redistribute it from -- there is nothing in `simple_data` for a
crossmap to transform into it. Apply the transform to the data the crossmap
*does* cover, then attach the new category's value afterwards:

```{r}
#| label: add-new-key

transformed <- apply_xmap(
  simple_data,
  simple_xmap,
  values_from = xcode_mass,
  keys_from = xcode
)

new_category <- tibble::tibble(alphacode = "F9", xcode_mass = 40)

bind_rows(transformed, new_category)
```

In both cases, the crossmap transform itself stays a pure redistribution
step -- every `xcode_mass` going in is accounted for coming out, nothing
more and nothing less -- and the actual judgment call about which categories
exist this period is made explicitly, outside `apply_xmap()`, where it's
visible rather than implicit.
