---
title: "Examining Collections of Crossmaps"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Examining Collections of Crossmaps}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: knitr-opts
#| include: false

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
#| label: setup
#| message: false

library(xmap)
library(dplyr)
library(stringr)
library(ggplot2)
```

### PLACEHOLDER: Re-extraction

```{r}
#| label: split-isic

split_isiccomb <- function(threefour_df) {
  interim <- list()
  interim$isiccomb.rows <-
    threefour_df |>
    filter(str_detect(isiccomb, "[:alpha:]"))
  interim$isiccomb.avg <-
    interim$isiccomb.rows |>
    group_by(country, year, isiccomb) |>
    mutate(value = tidyr::replace_na(value, 0)) |>
    summarise(
      avg.value = mean(value),
      n_isic = n_distinct(isic),
      n_rows = n(),
      .groups = "drop"
    )
  left_join(
    threefour_df,
    interim$isiccomb.avg,
    by = c("country", "year", "isiccomb")
  ) |>
    rename(value.nosplit = value) |>
    mutate(
      value = coalesce(avg.value, value.nosplit),
      split.isiccomb = !is.na(avg.value)
    ) |>
    select(country, year, isic, isiccomb, value, value.nosplit, split.isiccomb)
}

split_links <- indstat$masked_sample |>
  split_isiccomb() |>
  mutate(weights = value / 1000) |>
  tidyr::drop_na(weights)
```

As in `vignette("extract-validate-existing")`, we can confirm every
`country`/`year` group's links form a valid crossmap with `validate_as_xmap()`:

```{r}
#| label: validate-crossmaps

split_links |>
  nest_by(country, year, .key = "links") |>
  mutate(valid = validate_as_xmap(links, isiccomb, isic, weights)) |>
  ungroup() |>
  count(valid)
```

## Block-Level Imputation

The crossmaps framework addresses block-level transformation and imputation,
whereby instead of modifying rows of data, we treat the collection of code-value
pairs for a larger grouping entity as a single observation. For example, the
`isiccomb` code-value pairs in `vignette("extract-validate-existing")` are parts
of the overall aggregate output measure reported for a given country, year. The
extracted `isiccomb` to `isic` mapping is then a describe how to impute country,
year output measures by `isic` codes from the reported `isiccomb` observations.
The separation of this logic into a standalone object, the crossmap, allows for
the imputation to be examined in more granular detail compared to bespoke
scripts. This vignette illustrates some of the additional insights that can be
derived from analysing crossmaps.

We begin with the collection of extracted country, year `isiccomb` to `isic`
crossmaps from `vignette("extract-validate-existing")`. In that vignette, we
confirmed that the extracted weights formed valid crossmaps, guaranteeing that
the transformations preserved the overall total value. However, this does not
provide much insight into how different country, year observations are
transformed. In particular, not every country, year observation will be imputed,
many observations are already reported directly in `isic` categories. To
understand how much of the larger dataset and which country, year observation
blocks require imputation, we can examine properties of the crossmaps directly.
To begin, we might look at the distribution of weight types (unit, or fraction)
for each country, year:

```{r}
#| label: fig-weight-type

weight_type <- split_links |>
  dplyr::left_join(
    indstat$country_lookup[c("code", "iso3c", "name")],
    by = join_by(country == code)
  ) |>
  rename(country_iso3c = "iso3c") |>
  mutate(weight_type = ifelse(weights == 1, "unit", "frac"))

weight_type |>
  ggplot(aes(x = year, fill = weight_type)) +
  facet_wrap(~country_iso3c) +
  geom_bar(position = "stack") +
  scale_fill_manual(values = c("frac" = "#ef8a62", "unit" = "#67a9cf")) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  labs(y = "`isic` codes", x = "Year", fill = "Weight type")
```

The chart above shows us that for most country, year observations, fractional or splits affect relatively few imputed `isic` code-value pairs, with most values already directly reported in `isic` codes. Notice that for most country, years, combined reporting only affects a small number of final `isic`. Both USA and China (CHN) always reported directly in the `isic` classification, while Germany (DEU) mostly reported combined `isiccomb` values over a subset of `isic` values up from 1991 to 1994 inclusive. Yemen (YEM) is the conspicuous exception, where fractional weights dominate. We can summarise this in a fraction measuring how many `isic` code-value pairs are derived from splitting `isiccomb` values:

```{r}
#| label: fig-pct-isiccomb-heatmap

group_summary <- weight_type |>
  summarise(
    any_isiccomb = any(weight_type == "frac"),
    pct_isiccomb = sum(weight_type == "frac") / n(),
    .by = c(country_iso3c, year)
  ) |>
  mutate(pct_year_isiccomb = sum(any_isiccomb) / n(), .by = c(country_iso3c))

group_summary |>
  ggplot(aes(
    x = as.factor(year),
    fill = pct_isiccomb,
    y = forcats::fct_reorder(country_iso3c, pct_year_isiccomb)
  )) +
  geom_tile(color = "black") +
  ## abbreviate year labels
  scale_x_discrete(labels = ~ str_sub(.x, -2)) +
  scale_fill_stepsn(
    colours = c("white", RColorBrewer::brewer.pal(6, "Oranges")),
    breaks = c(0.0001, seq(0.2, 1, 0.2)),
    limits = c(0, 1),
    labels = scales::label_number(accuracy = 0.01)
  ) +
  # theme_minimal() +
  # move legend to bottom
  theme(legend.position = "bottom") +
  labs(
    x = "Year",
    y = "Country",
    fill = "Fraction of `isic` derived from `isiccomb` split",
    title = "Country, Year crossmaps for splitting reported values across combinations of ISIC Rev. 3 codes",
    subtitle = "Sorted by no. of years with at least one `isic` value derived from an `isiccomb` code"
  )
```

We can also create a binary indicator `any_isiccomb` for a more coarse-grain overview of which country, year have output values that are redistributed from an actual reported observation under a `isiccomb` code. This gives an overview of which parts of the harmonised `isic` dataset have some degree of imputation (orange tiles), and which remain unmodified from the reported values (blue tiles).

```{r}
#| label: fig-any-isiccomb-heatmap

group_summary |>
  ggplot(aes(
    x = as.factor(year),
    y = forcats::fct_reorder(country_iso3c, pct_year_isiccomb),
    fill = any_isiccomb
  )) +
  geom_tile(color = "black") +
  scale_x_discrete(labels = ~ str_sub(.x, -2)) +
  scale_fill_manual(values = c("TRUE" = "#ef8a62", "FALSE" = "#67a9cf")) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  labs(
    x = "Year",
    y = "Country",
    fill = "Crossmap involves at least 1 `isiccomb` split",
    title = "Country, Year crossmaps for splitting reported values across combinations of ISIC Rev. 3 codes",
    subtitle = "Sorted by no. of years with at least one `isic` value derived from an `isiccomb` code"
  )
```

## Hierarchical classifcations: Composing to coarser classes

The International Standard Industrial Classification is a strictly hierarchical classification system, whereby each code belongs to one of four nested levels: a 1-letter section, a 2-digit division, a 3-digit group, and a 4-digit class. A class's first three digits give its group parent, and (since `indstat` only reports at 3- or 4-digit granularity) `isic` codes in this dataset are either groups or classes.

Imagine that for analysis purposes we decide on using the 3-digit view of the data.  we could easily derive this by calculating the variable `isic3 = str_sub(isic, 1, 3)` and summarising the 4-digit values within each three digit group with a `sum()` operation. However, instantiating this aggregation as an explicit crossmap, surfaces some structure in the data that the more efficient aggregation would hide. To illustrate the additional insights we gain, we first create the `isic -> isic3` crossmap with `as_xmap_tbl()`, which also (trivially) validates the links. Unlike `isiccomb -> isic` -- it doesn't vary by `country` or `year`:

```{r}
#| label: isic3-xmap

(isic3_xmap <- split_links |>
  distinct(isic) |>
  mutate(isic = as.character(isic), isic3 = str_sub(isic, 1, 3), weight = 1) |>
  as_xmap_tbl(from = isic, to = isic3, weight_by = weight))
```

The resulting crossmap `isic3_xmap` takes 151 `isic` 3 and 4 digit codes and maps them to 61 `isic3` 3-digit codes. Given that the target set `.to$isic3 ` is smaller than the source set `.from$isic`, there must be some aggregation, but how much aggregation is unclear. For example, `1520` is a 4-digit class, but the `0` end digit actually indicates that division `152` is not further decomposed. In ISIC the ending digits `1-9` are used to label legitimate 4-digit sub-classes. This means that the value of `152` is the same as `1520`. However, `1520` could be a partial value split from a larger `isiccomb` code.

To gain an overview of the relationship between ISIC three and four digit classes, we can summarise the crossmap table further using similar summary methods as in the Timor-Leste occupation codes example in `vignette("extract-validate-existing")`:

```{r}
#| label: isic3-summary

(isic3_summary <- isic3_xmap |>
  summarise(
    `.from` = glue::glue_collapse(.from, "+"),
    `n.from` = n(),
    .by = c(.to)
  ) |>
  arrange(desc(n.from)))
```

The summary shows that for any given country, year observation with four-digit data, the largest number of four-digit values that could be aggregated is 9 values into the three-digit code `292`. A simple facetted heatmap shows the distribution more clearly:

```{r}
#| label: fig-isic3-waffle

isic3_agg <- isic3_summary |>
  mutate(agg = n.from > 1)

isic3_agg |>
  tidyr::uncount(n.from, .id = "unit", .remove = FALSE) |>
  mutate(agg = forcats::fct_relevel(factor(agg), "TRUE", "FALSE")) |>
  ggplot(aes(
    y = forcats::fct_reorder(.to$isic3, n.from),
    x = unit,
    fill = n.from,
  )) +
  facet_wrap(vars(agg), scales = "free_y") +
  geom_tile(width = 0.8, height = 0.8, color = "white") +
  scale_x_continuous(breaks = scales::breaks_width(1)) +
  scale_fill_viridis_c(breaks = scales::breaks_width(1)) + # + coord_equal()
  labs(
    x = "isic code (ordinal position within isic3 group)",
    y = "isic3 code",
    fill = "no. of isic codes aggregated",
    title = "Number of isic codes aggregated into each isic3 code",
    subtitle = "Faceted by whether the isic3 code aggregates more than one isic code"
  )
```

## Composing crossmap sequences: splitting then aggregating

`compose_xmap()` chains that shared `isic -> isic3` map with each group's own
`isiccomb -> isic` crossmap, producing `isiccomb -> isic3` directly -- without
ever materialising the intermediate `isic`-level values:

```{r}
#| label: isiccomb-to-isic3

isiccomb_to_isic3 <- split_links |>
  mutate(isic = as.character(isic), isiccomb = as.character(isiccomb)) |>
  nest_by(country, year, .key = "links") |>
  mutate(
    xmap1 = list(as_xmap_tbl(links, isiccomb, isic, weights)),
    composed = list(tidyr::unpack(
      compose_xmap(xmap1, isic3_xmap),
      everything()
    ))
  ) |>
  reframe(composed)

isiccomb_to_isic3 |>
  filter(country == "276", year == 1991, isiccomb == "151A")
```

`"151A"` still lands 1/5 of its weight on each of `151, 152, 153, 154, 155` --
`1520`'s composed weight rolls straight onto `isic3 = 152` -- the same picture a
two-step split-then-reaggregate would give, just arrived at directly.

### Reconvergent vs. crossing splits

Composing to `isic3` directly, rather than reaggregating by hand, makes a
structural property of each `isiccomb` group visible: does the group's
redistribution stay within a single `isic3` parent (**reconvergent** -- the
split only ever affected which 4-digit child got credited, and rolling up to 3
digits undoes it exactly), or does it span more than one `isic3` parent
(**crossing** -- the redistribution genuinely reallocates value between
different 3-digit categories, and that uncertainty survives the roll-up)?

```{r}
#| label: crossing-by-country

crossing_by_country <- split_links |>
  mutate(isic3 = str_sub(as.character(isic), 1, 3)) |>
  group_by(country, year, isiccomb) |>
  mutate(crosses_isic3 = n_distinct(isic3) > 1) |>
  ungroup() |>
  left_join(
    indstat$country_lookup[c("code", "iso3c")],
    by = c("country" = "code")
  ) |>
  group_by(iso3c) |>
  summarise(pct_crossing = mean(crosses_isic3) * 100, .groups = "drop") |>
  arrange(desc(pct_crossing))

crossing_by_country
```

This is exactly why Colombia and Yemen are in this sample: Colombia's `isiccomb`
splits are entirely reconvergent (0% cross a 3-digit boundary) -- every one is
fully recoverable once aggregated to `isic3`, since only the intra-group
allocation across 4-digit children was ever in question. Yemen sits at the other
extreme: \~95% of its split `isic` values belong to `isiccomb` groups that cross
a 3-digit boundary, so the naive even split propagates real allocation
uncertainty all the way into the 3-digit totals -- rolling up to `isic3` doesn't
make that uncertainty go away.

Applying the composed crossmap to the *original* `isiccomb`-level reported
values reproduces the same 3-digit totals as splitting to `isic` first and
reaggregating afterwards:

```{r}
#| label: direct-isic3

comb_values <- indstat$masked_sample |>
  mutate(isiccomb = as.character(isiccomb)) |>
  group_by(country, year, isiccomb) |>
  summarise(value = value[!is.na(value)][1], .groups = "drop") |>
  filter(!is.na(value))

direct_isic3 <- comb_values |>
  nest_by(country, year, .key = "values") |>
  mutate(
    xmap_grp = list(
      isiccomb_to_isic3 |>
        filter(.data$country == .env$country, .data$year == .env$year) |>
        as_xmap_tbl(isiccomb, isic3, weight_by)
    ),
    applied = list(apply_xmap(
      values,
      xmap_grp,
      values_from = value,
      keys_from = isiccomb
    ))
  ) |>
  reframe(applied)

direct_isic3
```

Comparing this against the two-step route -- `split_links`' `isic`-level values,
reaggregated up to 3 digits by hand -- confirms the two are identical:

```{r}
#| label: two-step-isic3-compare

two_step_isic3 <- split_links |>
  mutate(isic3 = str_sub(as.character(isic), 1, 3)) |>
  group_by(country, year, isic3) |>
  summarise(value_two_step = sum(value, na.rm = TRUE), .groups = "drop")

direct_isic3 |>
  full_join(two_step_isic3, by = c("country", "year", "isic3")) |>
  mutate(diff = value - value_two_step) |>
  summarise(max_abs_diff = max(abs(diff), na.rm = TRUE))
```

This only confirms the composition is mechanically correct, not that the
underlying naive split is accurate -- `indstat$masked_sample` replaces every
real reported value with `1000`, so it can't demonstrate how far off an even
split actually is from the true distribution (that's exactly why the
reconvergent/crossing distinction above has to come from the code structure
itself, not the values). Whether -- and how much -- that naive split matters for
a crossing group like Yemen's is the reconciliation problem tracked separately
in [#28](https://github.com/cynthiahqy/xmap/issues/28): preferring
directly-observed values over split ones wherever both exist, rather than
trusting the naive split everywhere.
