---
title: "Extracting Crossmaps from Existing Scripts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Extracting Crossmaps from Existing Scripts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE}
library(xmap)
library(dplyr)
```

## Case 1: Extracting logic as matrix

## Motivation

Many existing data harmonisation pipelines encode a crossmap's mapping logic implicitly, buried inside a legacy data-preparation script rather than represented as an explicit, checkable table. A common example is a Stata `do`-file that recodes a detailed source classification into a smaller set of target categories using a long sequence of `gen`/`replace` statements, e.g.:

```stata
gen farmer=0
replace farmer=1 if occupn>6000 & occupn<7000
gen teacher=0
replace teacher=1 if occupn>2400 & occupn<2500
gen professional=0
replace professional=1 if occupn>2000 & occupn<3000 & teacher==0
gen manager=0
replace manager=1 if occupn>1000 & occupn<1129
replace manager=1 if occupn>1131 & occupn<2000
gen armforces=0
replace armforces=1 if occupn<200
gen xefe=0
replace xefe=1 if occupn==1130
gen assprofclerk=0
replace assprofclerk=1 if occupn>3000 & occupn<5000
gen svcsales=0
replace svcsales=1 if occupn>5000 & occupn<6000
replace svcsales=1 if occupn>9000 & occupn<9200
gen labourer=0
replace labourer=1 if occupn>9200 & occupn<9320
gen driver=0
replace driver=1 if occupn>8320 & occupn<8330
replace driver=1 if occupn>9330 & occupn<9340
gen craftrademach=0
replace craftrademach=1 if occupn>7000 & occupn<9000 & driver==0
gen notclass=0
replace notclass=1 if occupn>9990 & occupn<10000
```

This kind of script is hard to audit: the mapping logic is scattered across many conditional statements, and there is no single artefact that documents which source codes map to which target category. This vignette walks through **extracting** the mapping logic implied by such a script into an explicit, validated `xmap_tbl`, without needing to touch the original (possibly sensitive) data at all.

The pipeline this vignette is drawn from keeps each step in its own script, tracing the actual `.do` file via Stata and reading the result back into R with `haven::read_dta()`. Here, the same extraction strategy is shown in full using only R, by writing a direct line-by-line R translation of the `.do` file's `gen`/`replace` logic in place of that trace step.

## Create a "carbon copy"

The key idea is that the recoding logic in a script like the one above only depends on the source *codes*, not on any other variables in the dataset. So instead of running the script on the real data, we can run it on a **"carbon copy"**: a placeholder dataset containing nothing but the distinct source codes that appear in the real data.

```{r carbon-copy}
## stand-in for `distinct(occupation_data, occupn) |> arrange(occupn)`
carbon_copy <- tibble::tibble(
  occupn = c(
    110, 120, 1110, 1120, 1130, 2111, 2112,
    2410, 2421, 3111, 3112, 5111, 5112,
    6111, 6112, 7111, 7112, 8321, 8322,
    9211, 9212, 9998
  )
)
```

## Trace script output with the carbon copy

Passing `carbon_copy` through the original script — in place of the real data — produces the same recoded output that running it on the real data would, but without ever exposing individual records. Below is a direct R translation of the `.do` file above: each `gen x = 0` becomes a new `0L` column, and each `replace x = 1 if <condition>` becomes a subset-assignment using the same condition, in the same order:

```{r r-equivalent-do-file}
## R equivalent of the `.do` file above
recode_occupn <- function(occupn) {
    df <- tibble::tibble(occupn = occupn)

    df$farmer <- 0L
    df$farmer[df$occupn > 6000 & df$occupn < 7000] <- 1L

    df$teacher <- 0L
    df$teacher[df$occupn > 2400 & df$occupn < 2500] <- 1L

    df$professional <- 0L
    df$professional[df$occupn > 2000 & df$occupn < 3000 & df$teacher == 0] <- 1L

    df$manager <- 0L
    df$manager[df$occupn > 1000 & df$occupn < 1129] <- 1L
    df$manager[df$occupn > 1131 & df$occupn < 2000] <- 1L

    df$armforces <- 0L
    df$armforces[df$occupn < 200] <- 1L

    df$xefe <- 0L
    df$xefe[df$occupn == 1130] <- 1L

    df$assprofclerk <- 0L
    df$assprofclerk[df$occupn > 3000 & df$occupn < 5000] <- 1L

    df$svcsales <- 0L
    df$svcsales[df$occupn > 5000 & df$occupn < 6000] <- 1L
    df$svcsales[df$occupn > 9000 & df$occupn < 9200] <- 1L

    df$labourer <- 0L
    df$labourer[df$occupn > 9200 & df$occupn < 9320] <- 1L

    df$driver <- 0L
    df$driver[df$occupn > 8320 & df$occupn < 8330] <- 1L
    df$driver[df$occupn > 9330 & df$occupn < 9340] <- 1L

    df$craftrademach <- 0L
    df$craftrademach[df$occupn > 7000 & df$occupn < 9000 & df$driver == 0] <- 1L

    df$notclass <- 0L
    df$notclass[df$occupn > 9990 & df$occupn < 10000] <- 1L

    df
}

(inc_df <- recode_occupn(carbon_copy$occupn))
```

The result is exactly the shape a Stata `.do` file of this kind would `save` as a `.dta` file, and that `haven::read_dta()` would read back into R: one row per source code, one dummy column per target category. This is an **incidence matrix**, just stored as a data frame with an `occupn` key column rather than as row names:

```{r inc-matrix}
inc_mtx <- as.matrix(inc_df[-1])
rownames(inc_mtx) <- inc_df[[1]]

inc_mtx[1:6, 1:6]
```

## Validate and summarise the crossmap

An incidence matrix is a wide representation of the same information a crossmap link list holds: which source key relates to which target key, and with what weight. `as_xmap_tbl()` expects a long `from`/`to`/`weights` table, so the matrix is pivoted to long form and the zero (i.e. unrelated) entries are dropped. This kind of matrix-to-links conversion isn't (yet) exported by `xmap`, so we define it inline:

```{r links-from-matrix}
## TODO: add fnc to the xmap package
links_from_matrix <- function(inc_mtx, rownames = "from") {
    stopifnot(is.matrix(inc_mtx))
    stopifnot(!is.null(rownames(inc_mtx)))
    stopifnot(!is.null(colnames(inc_mtx)))

    inc_mtx |>
        tibble::as_tibble(rownames = rownames) |>
        tidyr::pivot_longer(
            cols = !any_of(rownames),
            values_to = "weights",
            names_to = "to"
        ) |>
        dplyr::filter(weights != 0)
}

(occupn_links <- links_from_matrix(inc_mtx))
```

The extracted links can now be checked and coerced into a validated `xmap_tbl`, exactly as with links specified by hand:

```{r diagnose}
xmap::diagnose_as_xmap_tbl(occupn_links, from, to, weights)
```

```{r as-xmap}
(occupn_xmap <- occupn_links |>
    xmap::as_xmap_tbl(from, to, weights))
```

Because the recoding script only ever assigns a source code to a single target category, every link has a unit weight: this is a many-to-one aggregation, not a redistribution. Summarising by target category recovers, explicitly, the same grouping that was previously implicit in the script's `if` conditions — which source codes were collapsed into each target category:

```{r summary-by-target}
## not (yet) exported by the xmap package
summary_by_target <- function(x, from, to, weights, ...,
                               parts_into = "parts",
                               collapse = "+") {
    x |>
        dplyr::group_by({{ to }}) |>
        dplyr::summarise("{parts_into}" := glue::glue_collapse({{ from }}, collapse))
}

(summary_to_target <- occupn_links |>
    summary_by_target(from, to, weights, collapse = ","))
```

This crossmap can now be applied to real source-classification data with `apply_xmap()`, checked into version control as a documented artefact, or compared against a second, independently derived crossmap for the same recoding — such as the original Stata `.do` file traced on the same carbon copy — to check the two agree.
