---
title: "Extracting Crossmaps from Existing Scripts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Extracting Crossmaps from Existing Scripts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE}
library(xmap)
library(dplyr)
```

## Motivation

Many existing data harmonisation pipelines encode a crossmap's mapping logic implicitly, buried inside a legacy data-preparation script rather than represented as an explicit, checkable table. This kind of script is hard to audit: the mapping logic is scattered across many conditional statements, and there is no single artefact that documents which source codes map to which target category. This vignette walks through **extracting** the mapping logic implied by such a script into an explicit, validated `xmap_tbl`, without needing to touch the original (possibly sensitive) data at all.

## Case 1: Recoding & Aggregation

### The opaque recoding script

Imagine you are given an existing project with a data harmonisation script:

```stata
use "C:\Users\Folder\input.dta", clear

gen farmer=0
replace farmer=1 if occupn>6000 & occupn<7000
gen teacher=0
replace teacher=1 if occupn>2400 & occupn<2500
gen professional=0
replace professional=1 if occupn>2000 & occupn<3000 & teacher==0
gen manager=0
replace manager=1 if occupn>1000 & occupn<1129
replace manager=1 if occupn>1131 & occupn<2000
gen armforces=0
replace armforces=1 if occupn<200
gen xefe=0
replace xefe=1 if occupn==1130
gen assprofclerk=0
replace assprofclerk=1 if occupn>3000 & occupn<5000
gen svcsales=0
replace svcsales=1 if occupn>5000 & occupn<6000
replace svcsales=1 if occupn>9000 & occupn<9200
gen labourer=0
replace labourer=1 if occupn>9200 & occupn<9320
gen driver=0
replace driver=1 if occupn>8320 & occupn<8330
replace driver=1 if occupn>9330 & occupn<9340
gen craftrademach=0
replace craftrademach=1 if occupn>7000 & occupn<9000 & driver==0
gen notclass=0
replace notclass=1 if occupn>9990 & occupn<10000
```

You have access to the input data. Note the harmonisation index variable `occupn`:

```{r}
xmap::timor_occupn
```


You are unable to run the original STATA script, but would like to reproduce the harmonised data in R, and examine the applied mappings. You are interested in knowing:

- if any `occupn` codes in the `timor_occupn` data were missed in the harmonisation process, leading to silent loss of observations
- how many and which original `occupn` codes were mapped to each replacement occupation (e.g. `farmer`, `teacher`, etc.)

These questions are not possible to answer from the input and output data alone, and it should be clear that parsing the script itself is quite difficult.

### Recovering the mapping via carbon-paper substitution

The crossmaps framework offers a clear approach for extracting harmonisation logic from scripts. The basic idea is to pass "simplified data" through the script, and identify the mapping relationships based on the output. The simplified data can be thought of as 'carbon paper'. To demonstrate this extraction process in this vignette, we first rewrite the logic from the STATA script in R (using a LLM):

```{r r-equivalent-do-file}
recode_occupn <- function(occupn) {
    df <- tibble::tibble(occupn = occupn)

    df$farmer <- 0L
    df$farmer[df$occupn > 6000 & df$occupn < 7000] <- 1L

    df$teacher <- 0L
    df$teacher[df$occupn > 2400 & df$occupn < 2500] <- 1L

    df$professional <- 0L
    df$professional[df$occupn > 2000 & df$occupn < 3000 & df$teacher == 0] <- 1L

    df$manager <- 0L
    df$manager[df$occupn > 1000 & df$occupn < 1129] <- 1L
    df$manager[df$occupn > 1131 & df$occupn < 2000] <- 1L

    df$armforces <- 0L
    df$armforces[df$occupn < 200] <- 1L

    df$xefe <- 0L
    df$xefe[df$occupn == 1130] <- 1L

    df$assprofclerk <- 0L
    df$assprofclerk[df$occupn > 3000 & df$occupn < 5000] <- 1L

    df$svcsales <- 0L
    df$svcsales[df$occupn > 5000 & df$occupn < 6000] <- 1L
    df$svcsales[df$occupn > 9000 & df$occupn < 9200] <- 1L

    df$labourer <- 0L
    df$labourer[df$occupn > 9200 & df$occupn < 9320] <- 1L

    df$driver <- 0L
    df$driver[df$occupn > 8320 & df$occupn < 8330] <- 1L
    df$driver[df$occupn > 9330 & df$occupn < 9340] <- 1L

    df$craftrademach <- 0L
    df$craftrademach[df$occupn > 7000 & df$occupn < 9000 & df$driver == 0] <- 1L

    df$notclass <- 0L
    df$notclass[df$occupn > 9990 & df$occupn < 10000] <- 1L

    df
}
```

To form the 'simplified data' we extract the unique `occupn` codes from the `timor_occupn` data:

```{r}
(src_occupn <- unique(timor_occupn$occupn))
```

Then we can recover the transformation matrix by passing `src_occupn` into `recode_occupn()`:

```{r}
(out_df <- recode_occupn(src_occupn))
```

The matrix shown is an adjacency matrix showing binary connections between occupation codes in the original data, and the replacement codes specified by the script.

### Validating, building, and examining the crossmap

`out_df` already *is* an adjacency matrix — rows keyed by `occupn`, columns by replacement occupation, cells the (binary, here) weights — so we can check it's a valid crossmap directly with `validate_as_xmap()`. Trying that immediately surfaces a real data issue:

```{r, error = TRUE}
occupn_matrix <- out_df |>
  tibble::column_to_rownames("occupn") |>
  as.matrix()
```

Note that `out_df$occupn` includes `NA` because `timor_occupn` had observations whose original occupation code was never classified. We can confirm this by looking at the retrieved weights directly and noting there are no links -- i.e. all weights are 0:
```{r}
out_df |> filter(is.na(occupn))
```

However, a matrix row can't have a missing name, so `column_to_rownames()` aborts. We drop the row explicitly before validating the remaining links:

```{r}
occupn_matrix <- out_df |>
  tidyr::drop_na(occupn) |>
  tibble::column_to_rownames("occupn") |>
  as.matrix()

xmap::validate_as_xmap(occupn_matrix)
```

Now that `validate_as_xmap()` confirms `occupn_matrix` is a valid crossmap, we can coerce it directly into an `xmap_tbl` with `as_xmap_tbl()`'s matrix method — no reshape to long format needed, and unlinked pairs (weight = 0) are dropped automatically:

```{r as-xmap}
(occupn_xmap <- occupn_matrix |>
    xmap::as_xmap_tbl(from = "occupn", to = "replacement"))
```

From the crossmap tibble, we can see that 160 unique `occupn` codes are mapped into 12 `replacement` occupations. To understand the mapping further, we can summarise the weights to see that the original script only recodes and aggregates, but never splits an existing `occupn` code into multiple replacement codes.

Notice that distribution weights from the original `occupn` codes are `1`:

```{r}
occupn_xmap |> group_by(.weight_by) |> count()
```

And all but two `replacement` categories (`notclass` and `xefe`) are aggregations of original `occupn` codes:

```{r}
occupn_xmap |> group_by(.to) |> count()
```

Because the recoding script only ever assigns a source code to a single target category, every link has a unit weight: this is a many-to-one aggregation, not a redistribution. Summarising by target category recovers, explicitly, the same grouping that was previously implicit in the script's `if` conditions — which source codes were collapsed into each target category:

```{r summary-by-target}
occupn_xmap |>
  group_by(.to) |>
  summarise(`.from$occupn` = glue::glue_collapse(.from, "+"))
```

This crossmap can now be applied to real source-classification data with `apply_xmap()`, checked into version control as a documented artefact, or compared against a second, independently derived crossmap for the same recoding — such as the original Stata `.do` file traced on the same carbon copy — to check the two agree.

## Case 2: Splits

Now imagine a more complex recoding function, extracted from a larger data preparation pipeline:

```{r split-isic, attr.source='.numberLines'}
library(dplyr)
library(testthat)
library(stringr)
split_isiccomb <- function(threefour_df){
  #' Helper function to split isiccomb values across isic codes
  #' @param threefour_df df with 3/4 digit values across isic & isiccomb
  
  # make list for interim tables
  interim <- list()
  
  # extract rows with isiccomb codes
  interim$isiccomb.rows <- 
    threefour_df %>%
    filter(., str_detect(isiccomb, '[:alpha:]'))
  
  # test that we are not losing any data through spliting
  test_that("No `country,year` has more than one recorded `value` per `isiccomb` group", {
    rows_w_many_values_per_isiccomb <- 
      interim$isiccomb.rows %>%
      group_by(country, year, isiccomb) %>%
      ## get  no of recorded (not NA) values for given `country, year, isiccomb` 
      summarise(n_obs = sum(!is.na(value))) %>% 
      filter(n_obs != 1) %>%
      nrow()
    expect_true(rows_w_many_values_per_isiccomb == 0)
  })
  
  # calculate average value over isiccomb group for each country, year
  interim$isiccomb.avg <- 
    interim$isiccomb.rows %>%
    # group isiccomb rows, replace na with 0 for averaging
    group_by(country, year, isiccomb) %>%
    mutate(value = tidyr::replace_na(value,0)) %>%
    # split combination value over standard isic codes in isiccomb group
    summarise(avg.value = mean(value),
              ## checking variables
              n_isic = n_distinct(isic),
              n_rows = n()) %>%
    mutate(row_check = (n_isic == n_rows))
  
  #  return(interim$isiccomb.avg)
  
  ## check n_isic == n_rows
  test_that("isiccomb split average is calculated with correct denominator", {
    expect_true(all(interim$isiccomb.avg$row_check))
  })
  
  # output processed data
  final <-
    left_join(threefour_df, interim$isiccomb.avg, by = c('country', 'year', 'isiccomb')) %>%
    rename(value.nosplit = value) %>%
    mutate(value = coalesce(avg.value, value.nosplit),
           split.isiccomb = !is.na(avg.value)) %>%
    select(country, year, isic, isiccomb, value, value.nosplit, split.isiccomb) # not checking variables
  
  return(final)
}
```

You might start by asking an LLM to explain what this code does. For example, this explanation was generated by Claude Sonnet 5:

> This function handles a common problem in industrial classification data: source data sometimes reports a single value for a **combined** group of ISIC codes (e.g. a row where `isiccomb` = `"311/312"` covers two separate 3-digit ISIC classes) instead of reporting each code separately. `split_isiccomb()` disaggregates those combined values so every individual `isic` code ends up with its own value.
>
> Step by step:
>
> 1. **Identify combo rows** — filters to rows where `isiccomb` contains a letter (`str_detect(..., '[:alpha:]')`), which is presumably how combined codes are flagged (e.g. containing a suffix or separator letter rather than being purely numeric).
>
> 2. **Sanity check (test 1)** — for each `country, year, isiccomb` group, asserts there's exactly one non-`NA` `value` recorded. This guards against silently averaging/splitting when multiple conflicting values exist for the same combo — the split logic assumes one reported total per group.
> 
> 3. **Compute the split value** — groups by `country, year, isiccomb`, treats `NA` as 0, and takes `mean(value)` across the group. Since (per test 1) there's really only one non-zero value in the group, dividing by `n()` (the count of `isic` codes in that combo) effectively spreads the single reported total evenly across all the underlying ISIC codes — e.g. a value of 100 for `"311/312"` becomes 50 for `311` and 50 for `312`.
>
> 4. **Sanity check (test 2)** — asserts `n_isic == n_rows` per group, i.e. no duplicate `isic` codes within a combo group, so the denominator used for averaging is correct.
>
> 5. **Merge back** — left-joins the averaged/split values onto the original data, keeps both the original (`value.nosplit`) and split (`avg.value`) values, and uses `coalesce()` to prefer the split value where one exists, flagging which rows were split via `split.isiccomb`.
>
> In short: it's a data-cleaning helper that turns "one value shared across a combo of ISIC codes" into "one value per individual ISIC code," evenly divided, with two `testthat` guards to make sure that even split is actually valid for the data at hand.

Still, this explanation doesn't give much insight into which country, year observations contained combined values, and which ones didn't. In order to understand which country, year observations this split procedure was actually applied to, we can again extract the weights by using 'simplified data'. 

To keep things simple, we will extract weights for only 8 countries, across the following years where data was available. Five are large-economy reporters (USA, Japan, Germany, China and Brazil); the other three (Colombia, Romania and Yemen) are included because they exhibit structurally distinct splitting behaviour, which we return to below. The `indstat$country_lookup` table also carries an `income_group` column (the World Bank's 2006 historical classification), which spans all four groups across these 8 countries -- combined reporting is not a low-income-reporter phenomenon.

```{r}
library(ggplot2)
indstat$masked_sample |> 
  nest_by(country_iso3c, year) |>
  ggplot(aes(x = year, y = country_iso3c)) +
  geom_point()
```

We provide a masked version of the data, where the actual reported output values have been masked and replaced with the value `1000`. In the original data, the correspondence between `isic` and `isiccomb` is given in the same table as the reported output `value` for each `isiccomb` code, with duplicated rows for every `isic` code corresponding to a single `isiccomb` code. This can be seen in the observation `country=276`,`year=1991` shown below. The value for `151A` is `1000` (masked), and the code covers 5 `isic` codes (`151, 1520, 153, 154, 155`):

```{r}
indstat$masked_sample |>
  filter(country_iso3c == "DEU", year == 1991) |>
  filter(stringr::str_detect(isiccomb, '[:alpha:]'))
```

The structure of the original data further complicates understanding how the split-up `isic` values were calculated. Let's apply the transformation to the data to retrieve the mapping weights:

```{r}
split_links <- indstat$masked_sample |>
  split_isiccomb() |>
  mutate(weights = value/1000) |>
  tidyr::drop_na(weights)
```

We drop let's drop any unlinked combinations of `isic` and `isiccomb`, which will have `weight==NA` since no weights can be retrieved.

### Validating grouped links

`diagnose_as_xmap_tbl()` validates a single set of `.from`-`.to` links: it checks for duplicate pairs, missing weights, and that outgoing weights from each source sum to one. But `split_links` isn't a single crossmap — it's many crossmaps stacked together, one per `country`, `year`. The `isiccomb -> isic` mapping (and its weights) can differ from one `country`, `year` observation to the next, so validity has to be checked *within* each group, not across the whole table at once.

`diagnose_as_xmap_tbl()` always returns a single, stable `xmap_diagnosis` object — whether the crossmap is valid or not — so it composes directly with `dplyr::group_map()`, with no extra wrapper function needed. Applying it within each group of a grouped data frame, and collecting the results into a **nested tibble** (one row per group, the group's own links in a `data` list-column, a `valid` flag, and the group's `xmap_diagnosis` object in a `diagnosis` list-column) is a few lines of ordinary `dplyr`:

```{r}
group_diagnoses <- split_links |>
  group_by(country, year) |>
  group_map(\(group_df, group_key) {
    diagnosis <- diagnose_as_xmap_tbl(group_df, isiccomb, isic, weights)
    bind_cols(
      group_key,
      tibble::tibble(data = list(group_df), valid = diagnosis$valid, diagnosis = list(diagnosis))
    )
  }) |>
  bind_rows()

group_diagnoses |>
  count(valid)
```

### Understanding the transformation

Now that we've confirmed that the harmonisation links for each country, year are valid, we might be interested in understanding the harmonisation logic in greater detail. For this, let's look at the distribution of weight types for each country, year:


```{r}
library(ggplot2)
split_links |>
  dplyr::left_join(indstat$country_lookup[c("code", "iso3c", "name")], by = join_by(country == code)) |>
  rename(country_iso3c = "iso3c") |>
  mutate(weight_type = ifelse(weights == 1, "unit", "frac")) |>
  ggplot(aes(x = year, fill = weight_type)) +
  facet_wrap(~ country_iso3c) +
  geom_bar(position = "stack") +
  scale_fill_manual(values = c("frac" = "#ef8a62", "unit" = "#67a9cf")) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  labs(y = "`isic` codes", x = "Year", fill = "Weight type")
```

Notice that for most country, years, combined reporting only affects a small number of final `isic` -- Yemen is the conspicuous exception, where fractional weights dominate. We can also create a binary indicator `any_isiccomb` for a more coarse-grain overview of which country, year have output values that are redistributed from an actual reported observation under a `isiccomb` code. This gives an overview of which parts of the harmonised `isic` dataset have some degree of imputation, and which remain unmodified from the reported values.

```{r}
group_summary <- split_links |>
  mutate(weight_type = ifelse(weights == 1, "unit", "frac")) |>
  group_by(country, year) |>
  summarise(any_isiccomb = any(weight_type == "frac"))

group_summary |>
ggplot(aes(x = as.factor(year),
             fill = any_isiccomb,
             y = country
             )) +
    geom_tile(color = "black") +
    ## abbreviate year labels
    scale_x_discrete(labels = ~ str_sub(.x, -2)) +
    scale_fill_manual(values = c("TRUE" = "#ef8a62", "FALSE" = "#67a9cf")) +
    theme_minimal() +
    # move legend to bottom 
    theme(legend.position = "bottom") +
    labs(x = "Year",
        y = "Country",
        fill = "Split isiccomb codes?",
        title = "Countries reporting output values in combinations of ISIC Rev. 3 codes",
        subtitle = "INDSTAT 4 REV. 3 Output data")
```

### Does the split survive re-aggregation?

A fractional weight tells us a value was *imputed*, but not how much that imputation actually matters downstream. INDSTAT's 4-digit `isic` codes nest inside 3-digit ones (`isic.3`), and that `isic -> isic.3` aggregation is deterministic -- every `isic` code has exactly one `isic.3` parent, obtained by dropping the last digit. Composing the extracted `isiccomb -> isic` split with that aggregation gives an `isiccomb -> isic.3` crossmap, and at that coarser level a split can land in one of two very different places:

- **reconverges** -- all of the combo's `isic` children share a single `isic.3` parent, so the `1/n` weights sum back to `1` on re-aggregation. The split is imputed at 4 digits but *exact* at 3 digits and coarser.
- **crosses** -- the children fall under two or more `isic.3` parents, so the arbitrary even split propagates into the 3-digit totals as genuine allocation uncertainty.

Classifying each link only needs the `isic.3` parent and the extracted weights:

```{r}
link_types <- split_links |>
  mutate(isic.3 = str_sub(as.character(isic), 1, 3)) |>
  group_by(country, year, isiccomb) |>
  mutate(link_type = case_when(
    all(weights == 1)      ~ "unit",
    n_distinct(isic.3) == 1 ~ "reconverges",
    TRUE                    ~ "crosses"
  )) |>
  ungroup() |>
  left_join(indstat$country_lookup[c("code", "iso3c")], by = join_by(country == code)) |>
  rename(country_iso3c = "iso3c")

link_types |>
  count(country_iso3c, link_type) |>
  tidyr::pivot_wider(names_from = link_type, values_from = n, values_fill = 0)
```

Plotting the shares makes the spread across reporters obvious:

```{r}
ctry_order <- link_types |>
  group_by(country_iso3c) |>
  summarise(frac_crossing = mean(link_type == "crosses")) |>
  arrange(frac_crossing) |>
  pull(country_iso3c)

link_types |>
  mutate(link_type = factor(link_type, levels = c("unit", "reconverges", "crosses")),
         country_iso3c = factor(country_iso3c, levels = ctry_order)) |>
  ggplot(aes(y = country_iso3c, fill = link_type)) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = c("unit" = "#67a9cf", "reconverges" = "#a6dba0", "crosses" = "#ef8a62")) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  labs(x = "share of `isic` links", y = "Country", fill = NULL,
       title = "Do combined-reporting splits cross a 3-digit ISIC boundary?")
```

Three cases stand out. **Colombia** splits plenty of combined codes, but *none* of them cross a 3-digit boundary -- unlike China and the USA, which simply never split at all, Colombia splits *safely*: its 3-digit aggregates are exact. **Germany** sits in the middle, with about a tenth of its links crossing. **Yemen** is the extreme: some 95% of its `isic` links sit in crossing splits, so almost none of its 3-digit output totals are free of imputation. Romania shows the other structural extreme -- a single combo code, `2923A`, is split across six `isic` children that all reconverge into the `292` parent, sustained across 13 years:

```{r}
link_types |>
  filter(country_iso3c == "ROU", isiccomb == "2923A", year == 1990) |>
  select(year, isiccomb, isic, weights, link_type)
```

Once the `isiccomb -> isic` links are validated as a crossmap, `compose_xmap()` chains them with the deterministic `isic -> isic.3` crossmap into a single `isiccomb -> isic.3` crossmap, so 3-digit results can be produced without ever materialising the intermediate 4-digit values. That is exactly what makes the distinction above tractable: Romania's reconvergent weights cancel out inside the composition rather than being carried around as six imputed rows, and Yemen's crossing weights stay visible as explicit fractional links in the composed crossmap instead of being hidden in an already-aggregated total. See `vignette("xmap")` and `?compose_xmap` for details. For a fuller diagnostic treatment of reconverging vs. crossing structure across the whole INDSTAT extract, see [crossmap-explorer](https://github.com/cynthiahqy/crossmap-explorer).

### Invalid diagnosis

Let's modify one weights to create an invalid set of links.

```{r}
mod_links <- split_links
mod_links$weights[1] <- 0.7
```

Now, let's run the diagnosis again:

```{r}
invalid_diagnoses <- mod_links |>
  group_by(country, year) |>
  group_map(\(group_df, group_key) {
    diagnosis <- diagnose_as_xmap_tbl(group_df, isiccomb, isic, weights)
    bind_cols(
      group_key,
      tibble::tibble(data = list(group_df), valid = diagnosis$valid, diagnosis = list(diagnosis))
    )
  }) |>
  bind_rows()
```

Any invalid groups, and why they failed, can be inspected directly — each group's `diagnosis` entry prints a readable pass/fail report:

```{r}
bad_group <- invalid_diagnoses |>
  filter(!valid)

bad_group$diagnosis[[1]]

bad_group$data[[1]] |>
  filter(isiccomb == 151)
```