---
title: "Extracting Crossmaps from Existing Scripts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Extracting Crossmaps from Existing Scripts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE}
library(xmap)
library(dplyr)
```

## Motivation

Many existing data harmonisation pipelines encode a crossmap's mapping logic implicitly, buried inside a legacy data-preparation script rather than represented as an explicit, checkable table. This kind of script is hard to audit: the mapping logic is scattered across many conditional statements, and there is no single artefact that documents which source codes map to which target category. This vignette walks through **extracting** the mapping logic implied by such a script into an explicit, validated `xmap_tbl`, without needing to touch the original (possibly sensitive) data at all.

## Case 1: Extracting implicit matrix mappings

Imagine you are given an existing project with a data harmonisation script:

```stata
use "C:\Users\Folder\input.dta", clear

gen farmer=0
replace farmer=1 if occupn>6000 & occupn<7000
gen teacher=0
replace teacher=1 if occupn>2400 & occupn<2500
gen professional=0
replace professional=1 if occupn>2000 & occupn<3000 & teacher==0
gen manager=0
replace manager=1 if occupn>1000 & occupn<1129
replace manager=1 if occupn>1131 & occupn<2000
gen armforces=0
replace armforces=1 if occupn<200
gen xefe=0
replace xefe=1 if occupn==1130
gen assprofclerk=0
replace assprofclerk=1 if occupn>3000 & occupn<5000
gen svcsales=0
replace svcsales=1 if occupn>5000 & occupn<6000
replace svcsales=1 if occupn>9000 & occupn<9200
gen labourer=0
replace labourer=1 if occupn>9200 & occupn<9320
gen driver=0
replace driver=1 if occupn>8320 & occupn<8330
replace driver=1 if occupn>9330 & occupn<9340
gen craftrademach=0
replace craftrademach=1 if occupn>7000 & occupn<9000 & driver==0
gen notclass=0
replace notclass=1 if occupn>9990 & occupn<10000
```

You have access to the input data. Note the harmonisation index variable `occupn`:

```{r}
xmap::occupn_sample
```


You are unable to run the original STATA script, but would like to reproduce the harmonised data in R, and examine the applied mappings. You are interested in knowing:

- if any `occupn` codes in the `occupn_sample` data were missed in the harmonisation process, leading to silent loss of observations
- how many and which original `occupn` codes were mapped to each replacement occupation (e.g. `farmer`, `teacher`, etc.)

These questions are not possible to answer from the input and output data alone, and it should be clear that parsing the script itself is quite difficult.

## Extracting harmonisation logic

The crossmaps framework offers a clear approach for extracting harmonisation logic from scripts. The basic idea is to pass "simplified data" through the script, and identify the mapping relationships based on the output. The simplified data can be thought of as 'carbon paper'. To demonstrate this extraction process in this vignette, we first rewrite the logic from the STATA script in R:

```{r r-equivalent-do-file}
recode_occupn <- function(occupn) {
    df <- tibble::tibble(occupn = occupn)

    df$farmer <- 0L
    df$farmer[df$occupn > 6000 & df$occupn < 7000] <- 1L

    df$teacher <- 0L
    df$teacher[df$occupn > 2400 & df$occupn < 2500] <- 1L

    df$professional <- 0L
    df$professional[df$occupn > 2000 & df$occupn < 3000 & df$teacher == 0] <- 1L

    df$manager <- 0L
    df$manager[df$occupn > 1000 & df$occupn < 1129] <- 1L
    df$manager[df$occupn > 1131 & df$occupn < 2000] <- 1L

    df$armforces <- 0L
    df$armforces[df$occupn < 200] <- 1L

    df$xefe <- 0L
    df$xefe[df$occupn == 1130] <- 1L

    df$assprofclerk <- 0L
    df$assprofclerk[df$occupn > 3000 & df$occupn < 5000] <- 1L

    df$svcsales <- 0L
    df$svcsales[df$occupn > 5000 & df$occupn < 6000] <- 1L
    df$svcsales[df$occupn > 9000 & df$occupn < 9200] <- 1L

    df$labourer <- 0L
    df$labourer[df$occupn > 9200 & df$occupn < 9320] <- 1L

    df$driver <- 0L
    df$driver[df$occupn > 8320 & df$occupn < 8330] <- 1L
    df$driver[df$occupn > 9330 & df$occupn < 9340] <- 1L

    df$craftrademach <- 0L
    df$craftrademach[df$occupn > 7000 & df$occupn < 9000 & df$driver == 0] <- 1L

    df$notclass <- 0L
    df$notclass[df$occupn > 9990 & df$occupn < 10000] <- 1L

    df
}
```

To form the 'simplified data' we extract the unique `occupn` codes from the `occupn_sample` data:

```{r}
(src_occupn <- unique(occupn_sample$occupn))
```

Then we can recover the transformation matrix by passing `src_occupn` into `recode_occupn()`:

```{r}
(out_df <- recode_occupn(src_occupn))
```

The matrix shown is an adjacency matrix showing binary connections between occupation codes in the original data, and the replacement codes specified by the script.

## Validate and examine harmonisation mappings

To examine the extracted logic, we first transform the wide format matrix table into long format, dropping the unlinked pairs (weight = 0)

```{r}
library(tidyr)
links <- out_df |>
  tidyr::pivot_longer(
            cols = !any_of("occupn"),
            values_to = "weight",
            names_to = "replacement"
        ) |>
        dplyr::filter(weight != 0)
```

The extracted links can now be checked and coerced into a validated `xmap_tbl`, exactly as with links specified by hand:

```{r diagnose}
xmap::diagnose_as_xmap_tbl(links, from = occupn, to = replacement, weight = weight)
```

```{r as-xmap}
(occupn_xmap <- links |>
    xmap::as_xmap_tbl(from = occupn, to = replacement, weight = weight))
```

From the crossmap tibble, we can see that 160 unique `occpn` codes are mapped into 12 `replacement` occupations. To understand the mapping further, we can summarise the weights to see that the orginal script only recode and aggregates, but never splits an existing `occupn` code into multiple replacement codes.

Notice that distribution weights from the original `occupn` codes are `1`:

```{r}
occupn_xmap |> group_by(.weight_by) |> count()
```

And all but two `replacement` categories (`notclass` and `xefe`) are aggregations of original `occupn` codes:

```{r}
occupn_xmap |> group_by(.to) |> count()
```

Because the recoding script only ever assigns a source code to a single target category, every link has a unit weight: this is a many-to-one aggregation, not a redistribution. Summarising by target category recovers, explicitly, the same grouping that was previously implicit in the script's `if` conditions — which source codes were collapsed into each target category:

```{r summary-by-target}
occupn_xmap |>
  group_by(.to) |>
  summarise(`.from$occupn` = glue::glue_collapse(.from, "+"))
```

This crossmap can now be applied to real source-classification data with `apply_xmap()`, checked into version control as a documented artefact, or compared against a second, independently derived crossmap for the same recoding — such as the original Stata `.do` file traced on the same carbon copy — to check the two agree.

