---
title: "Handling Crossmap Tibbles"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{handling-xmap-tbls}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(xmap)
library(dplyr)
```


## Nesting and Invalidation of Crossmap Tibbles

In most cases, crossmap tibbles (`xmap_tbl`) should behave just like regular data frames or tibbles. However, we make use of nesting to retain meaningful variable names whilst also attaching `.from`, `.to` and `.weights` roles. Each column in an `xmap_tbl` actually contains a one-column tibble:

```{r nesting}
abc_xmap <- mock$abc_links |>
  as_xmap_tbl(lower, upper, share)
str(abc_xmap)
```

This nested structure can lead to unexpected behaviour when manipulating the `xmap_tbl` with standard `dplyr` verbs. This is somewhat intentional as subsetting can (silently) invalidate a crossmap (especially weights see `d -> CC` below):

```{r}
abc_xmap[1:4, ]
```

In most cases, we recommend flattening the crossmap tibble back to a standard tibble, modifying and then coercing it again back to a `xmap_tbl` to ensure weights are valid.

### Flattening and Exporting Crossmaps

<!-- This nesting adds some additional steps on top of standard the import and export operations for flat files. -->

There are a few ways to flatten or unpack a crossmap tibble. We recommend using `tidyr::unpack()` or `purrr:flatten_df()`, which both return tibbles:

```{r}
abc_xmap |>
  tidyr::unpack(dplyr::everything()) ## or

abc_xmap |>
  purrr::flatten_df()
```

When saving or exporting `xmap_tbl` objects as flat files (e.g. to `.csv`), you will need to first convert it into a standard tibble or data.frame without nesting.

```r
abc_xmap |>
  purrr::flatten_df() |>
  readr::write_csv("path/xmap.csv")
```

## Summarising Crossmaps

There are a number of features of crossmaps that might be of interest for documenting data provenance or preprocessing steps. We include here a selection of interesting properties and how to calculate them:

### Redistribution from Source Keys

If a crossmap involves any redistributions, `any(.xmap$.weight_by != 1)` will be true.
To find the links involved in redistribution:
```{r}
abc_xmap |>
  dplyr::filter(.weight_by[[1]] != 1)
```

### Composition of Target Keys

We can summarise which source keys contributed to each target:

```{r}
abc_xmap |>
  dplyr::group_by(.to) |>
  dplyr::summarise(".from({names(abc_xmap$.from)})" := paste(.from[[1]], collapse = ", "))
```

## Visualisation

Crossmap tibbles are valid edge lists, and can be visualised as graphs using packages such as [`ggraph`](https://ggraph.data-imaginist.com).